Further Evidence for Modified Spin-down in Sun-like Stars:\\Pile-ups in the Temperature--Period Distribution

### 2021-12-16 src/ms.tex
revising ages figure and discussion

\subsection{Ages of stars on the ridge}
\label{subsec:ages}
The WMB model predicts that hotter stars pile up on the long-period edge at younger ages, producing an age gradient across the edge. In Figure~\ref{fig:ages} we show the \teff-age distributions of stars on the ridge using ages from the CKS and SPOCS \citep{Brewer2018} catalogs, where we use the trapezoidal selection described in \S\ref{subsec:rossby} to select stars on the long-period pile-up. Both catalogs derive spectroscopic parameters from the CKS spectra, but use different pipelines for both the spectroscopic parameters and the isochrone fitting. In both cases, there appears to be an age gradient such that hotter stars are younger on average. However, such a trend is also expected in a sample of main-sequence stars as a natural consequence of the shorter main-sequence lifetimes of hotter, more massive stars. We also find that the dispersion in age is a sensitive function of \teff, with cooler stars on the ridge showing a broader range of ages. This observation could be due to cooler stars populating the ridge for longer periods of time (relative to hotter stars), the lower precision of isochrone ages for cooler stars, or some combination of the two effects.

We determined that 90\% of the stars on the ridge have ages between 1.4--5.6~Gyr (using ages from the CKS catalog), or 2.3--5.9~Gyr (using SPOCS ages). In the WMB model, although wind-driven angular momentum losses cease, stars continue to evolve structurally which results in evolution in the moment of inertia and stellar spin, driven by expansion of the stars away from the main-sequence \citep{vanSaders2019}. As a result, an age-gradient is expected along the pile-up as stars of different mass depart the pile-up at different ages. Though we have not measured this gradient, the data suggest that some stars spend several Gyr occupying the ridge with only modest evolution of their spin rates.

>>>>

\subsection{Ages of stars on the long-period pile-up}
\label{subsec:ages}
The WMB model predicts that hotter stars pile up on the long-period edge at younger ages, producing an age gradient across the edge. In Figure~\ref{fig:ages} we show the \teff-age distributions of stars on the ridge using ages from the CKS and SPOCS \citep{Brewer2018} catalogs, where we use the trapezoidal selection described in \S\ref{subsec:rossby} to select stars on the long-period pile-up. Both catalogs derive spectroscopic parameters from the CKS spectra, but use different pipelines for both the spectroscopic parameters and the isochrone fitting. In both cases, there appears to be an age gradient such that hotter stars are younger on average. However, such a trend is also expected in a sample of main-sequence stars as a natural consequence of the shorter main-sequence lifetimes of hotter, more massive stars. We also find that the dispersion in age is a sensitive function of \teff, with cooler stars on the ridge showing a broader range of ages. This observation could be due to cooler stars populating the ridge for longer periods of time (relative to hotter stars), the lower precision of isochrone ages for cooler stars, or some combination of the two effects.

We determined that 90\% of the stars on the ridge have ages between 1.4--5.6~Gyr (using ages from the CKS catalog), or 2.3--5.9~Gyr (using SPOCS ages). However, we note that systematic effects in surveys and theoretical models lead to large uncertainties in isochrone ages that are not necessarily represented by the reported age uncertainties. In fact, there are only a handful of long-period pile-up stars in the CKS sample with ages older than the age of the Sun, and almost all of the pile-up stars would be compatible with an age equal to or less than the Sun's given the large age uncertainties.

In the WMB model, although wind-driven angular momentum losses cease, stars continue to evolve structurally which results in evolution in the moment of inertia and stellar spin, driven by expansion of the stars away from the main-sequence \citep{vanSaders2019}. As a result, an age-gradient is expected along the pile-up as stars of different mass depart the pile-up at different ages. Though we have not measured this gradient, the data suggest that some stars spend several Gyr occupying the ridge with only modest evolution of their spin rates.
